---
title: "R Notebook"
output:
  html_notebook: default
  github_document:
    toc: true
    toc_depth: 2
---


```{r}
library(ggplot2)
library(ggmap)
library(readxl)
library(cowplot)
```

# Read station data with popsitions of each station

```{r}
data_positions <- read_excel("Data/STAR-WALK sites.xlsx")
```

## Our first plot
This uses the basic plot() function of R. It tells R to plot the numbers of 'lat' on the y axis, the numbers of 'long' on the x axis, and these numbers R can find in the data set we have called 'data_positions' 
```{r}
plot(lat~long, data_positions)
```

## Another test plot
We use `color = lake` inside the aes() parathesis to make on color per lake.
```{r}
ggplot(data_positions, aes(long, lat, color = lake)) +
  geom_point()
```

## Yet another test plot
```{r}
ggplot(data_positions, aes(long, lat, color = lake)) +
  geom_point() +
  facet_wrap(~lake, scales = "free")
```


# Example of one lake plot (Lake Lura)
## Create background map
```{r}
# Make 'google' map
# mydata <- subset(data_positions, lake == "Biogradsko")
# map <- qmap(location = c(19.5983, 42.898), zoom=16)
mydata <- subset(data_positions, lake == "Lura")
mymap <- qmap(location = c(20.19387, 41.78988), zoom=16)
mymap
```

## Show station positions
### Just positions
```{r}
# Make 'google' map
mymap + 
  geom_point(data = mydata, aes(Elongitude, Nlatitude), color = "black")
```

### Shown station names also
```{r}
# Make 'google' map
mymap + 
  geom_point(data = mydata, aes(Elongitude, Nlatitude), color = "black") +
  geom_text(data = mydata, aes(x = Elongitude + 0.0003, y = Nlatitude, label=site), hjust = "left", color = "black")
```

## Plot chemistry

### Get chemistry data into R's memory
The data are in an excel file, so we use the command 'read_excel' to read the data into R's memory.  Note that  
 + `skip = 2` means that we skip the two first lines of the Excel file (we want the variable names to be in our row number 1)  
 + `na = "n.d."` means that the cells with no data are marked "n.d.". Otherwise R will think this is some kind of "data". 
```{r}
data_chem <- read_excel("Data/summary of most important chemistry results.xlsx", skip = 2, na = c("","n.d."))
```

### Add columns for longitude and latitude to the data
```{r}
data_chem <- merge(data_chem, data_positions, by = c("lake","site"), all.x = TRUE, all.y = FALSE)
```

### Plot (first try)
```{r}
mymap + 
  geom_point(data = data_chem, aes(Elongitude, Nlatitude, color = water_TP_autumn))
```

### Plot (second try)
```{r}
mymap + 
  geom_point(data = data_chem, aes(Elongitude, Nlatitude, color = water_TP_autumn), size = 5)
```

### Plot (color scale based on all lakes)
'data_chem' has data for _all_ lakes, and the colours scale is adapted to that. But because 'mymap' sets the limits of the map we only see the points around this lake - the other lakes are outside the picture.   
```{r}
mymap + 
  geom_point(data = data_chem, aes(Elongitude, Nlatitude, color = water_TP_autumn), size = 5) +
  scale_color_distiller(palette = "RdYlGn")
```

### Plot (color scale based on this lake only)
To make the colour scale adjust to only the lake we show, we have to make a new data set (we call it 'data_chem_mylake') for only the data of this lake. 
```{r}
data_chem_mylake <- subset(data_chem, lake == "Lura")
mymap + 
  geom_point(data = data_chem_mylake, aes(Elongitude, Nlatitude, color = water_TP_autumn), size = 5) +
  scale_color_distiller(palette = "RdYlGn")
```

### More things we can do to improve our plot
#### Add a line around the points
We do this by using 'shape = 21'.
 + Note that we have to change 'color' and 'scale_color_distiller' into 'fill' and 'scale_fill_distiller', because for this shape, 'color' is only the color of the thin line around the circle.
```{r}
mymap + 
  geom_point(data = data_chem_mylake, aes(Elongitude, Nlatitude, fill = water_TP_autumn), shape = 21, size = 5) +
  scale_fill_distiller(palette = "RdYlGn")
```

#### Change shape
Here we change 'shape = 21' to 'shape = 22'. Run the command `?points` to see the different shapes we can use. 
```{r}
mymap + 
  geom_point(data = data_chem_mylake, aes(Elongitude, Nlatitude, fill = water_TP_autumn), shape = 22, size = 5) +
  scale_fill_distiller(palette = "RdYlGn")
```
#### Change title of the legend 
We do this by adding the title just after the first parenthesis in the scale definition `scale_fill_distiller`.  
Note that `\p` means "line shift", so `Total water P\nAutumn` prints "Total water", line shift,  "Autumn". 
```{r}
mymap + 
  geom_point(data = data_chem_mylake, aes(Elongitude, Nlatitude, fill = water_TP_autumn), shape = 22, size = 5) +
  scale_fill_distiller("Total water P\nAutumn", palette = "RdYlGn")
```
#### Add plot title 
```{r}
mymap + 
  geom_point(data = data_chem_mylake, aes(Elongitude, Nlatitude, fill = water_TP_autumn), shape = 22, size = 5) +
  scale_fill_distiller("Total water P\nAutumn", palette = "RdYlGn") + 
  labs(title = "Lake Lura, total phosphorus")
```


#### Also use size 
 + Added `size = water_TP_autumn` _inside_ the aes() paranthesis
 + Remember to delete the `size = 5` _outside_ the aes() paranthesis
 + We can also show a different variable using size 
```{r}
mymap + 
  geom_point(data = data_chem_mylake, aes(Elongitude, Nlatitude, fill = water_TP_autumn, size = water_TP_autumn), shape = 22) +
  scale_fill_distiller("Total water P\nAutumn", palette = "RdYlGn")
```

#### Plot a different variable (sediment total P)
 + Changed `water_TP_autumn` to `sed_TP_autumn`
 + Remember to also change any titles (here, legend title)
```{r}
mymap + 
  geom_point(data = data_chem_mylake, aes(Elongitude, Nlatitude, fill = sed_TP_autumn), shape = 22, size = 5) +
  scale_fill_distiller("Total sediment P\nAutumn", palette = "RdYlGn")
```



## Plot diatom data

### Get diatom data into R's memory
```{r}
data_diatoms <- read_excel("Data/summary of the most important diatom results.xlsx", na = c("","n.d."))
```

### Add columns for longitude and latitude to the data
```{r}
data_diatoms <- merge(data_diatoms, data_positions, by = c("lake","site"), all.x = TRUE, all.y = FALSE)
```

### Plot IPS as colours on map
```{r}
mymap + 
  geom_point(data = data_diatoms, aes(Elongitude, Nlatitude, color = IPS), size = 5) +
  scale_color_distiller(palette = "RdYlGn")
```

## Plot macroinvertebrate data

### Get macroinvertebrate data into R's memory
```{r}
data_invert <- read_excel("Data/summary of most important macroinvertebrate data.xlsx", sheet = "indices_traits", skip = 1, na = c("","n.d."))
```

### Add columns for longitude and latitude to the data
```{r}
data_invert <- merge(data_invert, data_positions, by = c("lake","site"), all.x = TRUE, all.y = FALSE)
```

### Plot IPS as colours on map
```{r}
mymap + 
  geom_point(data = data_invert, aes(Elongitude, Nlatitude, color = ASPT), size = 5) +
  scale_color_distiller(palette = "RdYlGn")
```

## Plot macrophyte data

### Get macrophyte data into R's memory
```{r}
data_macrophyte <- read_excel("Data/summary of most important macrophyte results.xlsx", skip = 1, na = c("","n.d."))
```

### Add columns for longitude and latitude to the data
```{r}
data_macrophyte <- merge(data_macrophyte, data_positions, by = c("lake","site"), all.x = TRUE, all.y = FALSE)
```

### Plot IPS as colours on map
```{r}
mymap + 
  geom_point(data = data_macrophyte, aes(Elongitude, Nlatitude, color = MI), size = 5) +
  scale_color_distiller(palette = "RdYlGn")
```

# Make plot of all lakes
## Mean position of each lake
```{r}
data_positions %>%
  group_by(lake) %>%
  summarise(long = mean(Elongitude), lat = mean(Nlatitude)) %>%
  as.data.frame()
```

## Create background maps
```{r}
mymap1 <- qmap(location = c(19.5983, 42.898), zoom=16)
mymap2 <- qmap(location = c(19.09116, 43.14598), zoom=16)
mymap3 <- qmap(location = c(20.19387, 41.78988), zoom=16)
mymap4 <- qmap(location = c(20.73327, 41.03089), zoom=11)
mymap5 <- qmap(location = c(20.95873, 40.90712), zoom=11)
mymap6 <- qmap(location = c(20.39183, 44.78354), zoom=14)
```

## Create a list of maps and arrange them
```{r, fig.width = 8, fig.height = 12}
map_list <- list(mymap1, mymap2, mymap3, mymap4, mymap5, mymap6)
plot_grid(plotlist = map_list, ncol = 2)
```


## Create a list of maps, add points, and arrange them
```{r}
map_list <- list(mymap1, mymap2, mymap3, mymap4, mymap5, mymap6)
for (i in 1:6){
  map_list[[i]] <- map_list[[i]] +
    geom_point(data = data_chem, aes(Elongitude, Nlatitude, fill = water_TP_autumn), pch = 21, size = 5) +
    scale_fill_distiller("Total water P\nAutumn", palette = "RdYlGn")
}
```

```{r, fig.width = 10, fig.height = 12, warning=FALSE}
plot_grid(plotlist = map_list, ncol = 2)
```





# Plot pie diagrams on map
```{r}
library(scatterpie)

d <- mydata
n <- nrow(d)
d$region <- factor(1:n)
d$A <- abs(rnorm(n, sd=1))
d$B <- abs(rnorm(n, sd=2))
d$C <- abs(rnorm(n, sd=3))
d$D <- abs(rnorm(n, sd=4))
d$radius <- abs(rnorm(n)*0.0002)
```



```{r}
ggplot() + geom_scatterpie(aes(x=Elongitude, y=Nlatitude, group=region), data=d, cols=LETTERS[1:4])
```
```{r}
ggplot() + geom_scatterpie(aes(x=Elongitude, y=Nlatitude, group=region, r = radius), data=d, cols=LETTERS[1:4])
```

```{r}
mymap + geom_scatterpie(aes(x=Elongitude, y=Nlatitude, group=region), data=d, cols=LETTERS[1:4])
```
```{r}
mymap + geom_scatterpie(aes(x=Elongitude, y=Nlatitude, group=region, r = radius), data=d, cols=LETTERS[1:4])
```

```{r}

```





